The Natural History of Child Signals of Need in Utila, Honduras: An Exploratory Study

### 2025-05-23 Paper.qmd
Changes made for knitting and plot making (mainly updated dictionaries).

@@

source("analysis.R")
#source("analysis2.R") # Generates many plots. Slow!
```

@@

From these count variables we created three composite signaling variables to be used as outcome variables. The *frequency of signaling* was the sum of the frequencies of child sadness, crying, and temper tantrums. *Maximum signal frequency* was the frequency of the most common of these three signals. *Signal cost* was the sum of the frequencies of these three signals weighted by their relative costliness (i.e., *Signal cost* = *frequency of sadness* + 2 $\times$ *frequency of crying* + 3 $\times$ *frequency of temper tantrums*). Although we treated *signal cost* as a continuous variable, we conceptualized low signal cost values (e.g., infrequent sadness or crying) as more likely to be minimal signals, and high signal cost values (e.g., frequent crying or temper tantrums) as more likely to be "costly" signals.

@@

utila_df |>
  dplyr::select(any_of(names(utila_df[names(longform_dict[1:23])]))) |>
  setNames(longform_dict[1:23]) |>
  datasummary_skim(fun_numeric = list(N = \(x) length(na.omit(x)), Range = \(x) str_glue("{min(x, na.rm=T)}-{max(x, na.rm=T)}"), Mean = Mean, SD = SD, Histogram = function(x) ""))
```

@@

The proportion of children perceived by caretakers as being more needy, as needy, or less needy than a caretaker's other children when holding continuous variables at their mean and factor variables at their mode. A: Effect of child age on the probability of relative neediness B: Effect of child sadness frequency on the probability of relative neediness. Outcome variable question wording: How much time, effort, and resources does this child need compared to your other children?
:::

@@

Contrary to expectations, immigration status, residence in Camponado, and primary language (Spanish or English as proxies for cultural differences), did not predict child signaling in our lasso regressions, despite the differences in wealth, job certainty, discrimination, and living conditions between those from the mainland and Utilians [@davis2014; @garcia2017; @garcia2018; @stonich2000]. These variables correlated with neighborhood quality, however, which was retained in our models, and lasso regression tends to retain only one of a set of correlated predictors [@zou2005].

@@

Caregiver responses to signals of need were generally positive or neutral, more so for situations involving injury and illness, and less so for situations involving family conflict, loss of privileges, and transgressions, which had greater proportions of negative responses (@fig-effect-last-signal-al-responses; @fig-effect-last-signal-response). These patterns probably reflect a combination of shared interests vs. conflicts of interest between child and caretaker, some of which might represent caretaker investment in some children more than others, and greater caretaker certainty about need in cases of injury and illness. Negative responses, e.g., punishment, can be used to change future child behavior that is costly to the caregiver (e.g., harming siblings or wasting resources), and possibly also to signal the costliness of providing the desired support [@hagen2024; @gaffney2022]. That said, our model predicts positive responses in 35% of cases involving transgressions, suggesting that children are successful in negotiating better outcomes at least some of the time.

>>>>

set.seed(51300)

@@

source("analysis.R")
source("analysis2.R") # Generates many plots. Slow!
```

@@

From these count variables we created two composite signaling variables. The *frequency of signaling* was the sum of the frequencies of child sadness, crying, and temper tantrums. *Signal cost* was the sum of the frequencies of these three signals weighted by their relative costliness (i.e., *Signal cost* = *frequency of sadness* + 2 $\times$ *frequency of crying* + 3 $\times$ *frequency of temper tantrums*). Although we treated *signal cost* as a continuous variable, we conceptualized low signal cost values (e.g., infrequent sadness or crying) as more likely to be minimal signals, and high signal cost values (e.g., frequent crying or temper tantrums) as more likely to be "costly" signals. For each child we also summed the frequencies of conflict with caretakers and signaling cost for all other children in the household, termed *Other children conflict* and *Other children signal cost*, respectively.

@@

utila_df2 |>
  dplyr::select(any_of(names(utila_df2[names(longform_dict[1:25])]))) |>
  setNames(longform_dict[1:25]) |>
  datasummary_skim(fun_numeric = list(N = \(x) length(na.omit(x)), Range = \(x) str_glue("{min(x, na.rm=T)}-{max(x, na.rm=T)}"), Mean = Mean, SD = SD, Histogram = function(x) ""))
```

@@

The proportion of children perceived by caretakers as being more needy, as needy, or less needy than a caretaker's other children when holding continuous variables at their mean and factor variables at their mode. a: Effect of child age on the probability of relative neediness. b: Effect of child sadness frequency on the probability of relative neediness. Outcome variable question wording: How much time, effort, and resources does this child need compared to your other children?
:::

@@

Contrary to expectations, immigration status, residence in Camponado, and primary language (Spanish or English as proxies for cultural differences), did not predict child signaling in our lasso regressions, despite the differences in wealth, job certainty, discrimination, and living conditions between those from the mainland and English speaking Utilians [@davis2014; @garcia2017; @garcia2018; @stonich2000]. These variables correlated with neighborhood quality, however, which was retained in our models, and lasso regression tends to retain only one of a set of correlated predictors [@zou2005]. Additionally, families from Camponado appeared to be willing to care for children outside their immediate families, which might have reduced children's needs.

@@

Keeping in mind possible biases, caregiver self-reported responses to signals of need were generally positive or neutral, more so for situations involving injury and illness, and less so for situations involving family conflict, loss of privileges, and transgressions, which had greater proportions of negative responses (@fig-effect-last-signal-al-responses; @fig-effect-last-signal-response). These patterns probably reflect a combination of shared interests vs. conflicts of interest between child and caretaker, some of which might represent caretaker investment in some children more than others, and greater caretaker certainty about need in cases of injury and illness. Negative responses, e.g., punishment, can be used to change future child behavior that is costly to the caregiver (e.g., harming siblings or wasting resources), and possibly also to signal the costliness of providing the desired support [@hagen2024; @gaffney2022]. That said, our model predicts positive responses in 35% of cases involving transgressions, suggesting that children are successful in negotiating better outcomes at least some of the time.
